---
title: "Exploratory Analysis of Measles Susceptibility Trends in Glasgow (1998–2012)"
execute:
  echo: false
  warning: false
  message: false
editor_options:
  chunk_output_type: console
format:
  pdf:
    number-sections: true
    latex-engine: pdflatex
    papersize: a4
    geometry: top=1in, bottom=1in, left=1in, right=1in
    fontsize: 11pt
    include-in-header: header.tex
---

```{r}
library(ggplot2)
library(dplyr)
library(readr)
library(knitr)
```

## Exploratory Data Analysis

Exploratory Data Analysis (EDA) was conducted to understand trends in measles susceptibility among pre-school children in Glasgow from 1998 to 2012. This includes examining year-to-year patterns, comparing susceptibility before and after 2004, and visualizing distributions to support further statistical analysis.

```{r}
 measles_data <- read.csv("C:/Users/Axnou/OneDrive - University of Glasgow/staticstics/stats level 3/Data analysis/Gruop Project/DAProject15.csv") 

measles_data <- measles_data %>%
  mutate(susceptibility_rate = Y / N,
         period = ifelse(Year >= 2004, "After 2004", "Before 2004"))
```

## Summary statistic

```{r}
mean_susceptibility <- measles_data %>%
  group_by(Year) %>%
  summarise(mean_susceptibility = mean(susceptibility_rate, na.rm = TRUE))

kable(mean_susceptibility,
      caption = "Mean Measles Susceptibility by Year",
      digits = 4,
      align = "c")
```

*Table 1 shows the mean measles susceptibility by year. No sharp decline is observed post-2004, indicating that vaccine hesitancy may have persisted after the Wakefield article's retraction. Any recovery in vaccination rates appears gradual.*

## Data visualisation

```{r}
#| fig-height: 3.5
#| fig-width: 5
#| fig-cap: "Boxplot of Measles Susceptibility Rates by Year in Glasgow (1998–2012)"


ggplot(measles_data, aes(x = factor(Year), y = susceptibility_rate)) +
  geom_boxplot(fill = "skyblue") +
  geom_vline(xintercept = which(levels(factor(measles_data$Year)) == "2004"), 
             linetype = "dashed", color = "red") +
  labs(title = "Distribution of Susceptibility Rates by Year",
       x = "Year", y = "Susceptibility Rate") +
  theme_minimal()

```

```{r}
#| fig-height: 3.5
#| fig-width: 5
#| fig-cap: "Mean Measles Susceptibility Before vs After 2004"
measles_data |>
  group_by(period) |>
  summarise(mean_susceptibility = mean(susceptibility_rate, na.rm = TRUE)) |>
  ggplot(aes(x = period, y = mean_susceptibility, fill = period)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Before 2004" = "steelblue", "After 2004" = "tomato")) +
  labs(title = "Mean Susceptibility: Before vs After 2004",
       x = "Period", y = "Mean Susceptibility Rate") +
  theme_minimal()
```

Figures 1 and 2 together show measles susceptibility trends across years and between periods. Figure 1 highlights yearly distributions, with the 2004 retraction marked, showing a peak in 2008 but no sharp drop post-2004. Figure 2 compares average susceptibility before and after 2004, revealing only a slight decrease. These patterns align with Table 1, suggesting vaccine hesitancy persisted beyond the article’s retraction.
