---
title: "EDA: Measles Susceptibility in Glasgow"
execute:
  echo: false
  warning: false
  message: false
editor_options:
  chunk_output_type: console
format:
  pdf:
    papersize: a4
    geometry: margin=1in
    number-sections: true
    toc: true
    keep-tex: true
    latex-engine: pdflatex
    fontsize: 11pt
---

```{r}
library(ggplot2)
library(dplyr)
library(readr) 
library(knitr)

```

## Exploratory Data Analysis

Exploratory Data Analysis (EDA) was conducted to understand trends in measles susceptibility among pre-school children in Glasgow from 1998 to 2012. This includes examining year-to-year patterns, comparing susceptibility before and after 2004, and visualizing distributions to support further statistical analysis.

```{r}

measles_data <- read.csv("C:/Users/Axnou/OneDrive - University of Glasgow/staticstics/stats level 3/Data analysis/Gruop Project/DAProject15.csv")

measles_data <- measles_data |> 
  mutate(susceptibility_rate = Y / N, 
         period = ifelse(Year >= 2004, "After 2004", "Before 2004"))


```

## Summary statistic

```{r}
 mean_susceptibility <- measles_data %>% 
   group_by(Year) |> 
   summarise(mean_susceptibility = mean(susceptibility_rate, na.rm = TRUE))

kable(mean_susceptibility, 
      caption = "Mean Measles Susceptibility by Year", 
      digits = 4, 
      align = "c")

```

*Table 1 shows the mean measles susceptibility by year. No sharp decline is observed post-2004, indicating that vaccine hesitancy may have persisted after the Wakefield article's retraction. Any recovery in vaccination rates appears gradual.*

## Data visualisation

```{r}
#| fig-cap: "Measles Susceptibility Trend in Glasgow (1998–2012)" 

ggplot(measles_data, aes(x = Year, y = susceptibility_rate)) +
  geom_point(size = 3, color = "steelblue") +
  geom_vline(xintercept = 2004, linetype = "dashed", color = "red") + 
  labs(title = "Measles Susceptibility Over Time", x = "Year", y = "Proportion Susceptible") +
  theme_minimal()
```

```{r}
#| fig-cap: "Mean Measles Susceptibility Before vs After 2004" 

measles_data %>% group_by(period) |>
  summarise(mean_susceptibility = mean(susceptibility_rate, na.rm = TRUE)) |>
  ggplot(aes(x = period, y = mean_susceptibility, fill = period)) + 
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Before 2004" = "steelblue", "After 2004" = "tomato")) + 
  labs(title = "Mean Susceptibility: Before vs After 2004", x = "Period", y = "Mean Susceptibility Rate") + 
  theme_minimal()
```

*Figures 1 and 2 show susceptibility trends over time and grouped by period. No immediate decline follows the 2004 retraction. This implies the article's retraction had limited short-term impact on susceptibility, reinforcing the persistence of vaccine hesitancy.*
